Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-06 paper/ms.tex
Added to discussion of applications.

There are two primary applications we imagine for the \genrp formalism: 1). modeling
astrophysical variability to infer the properties of the object that is varying; 2).
account for astrophysical variability as a source of noise when trying to detect 
additional phenomena; 3). interpolate or extrapolate variations to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of quasars, measuring the asteroseismic variations of stars,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in 
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

Our background is in studying transiting exoplanets, which has only recently begun
to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
One early successful attempt at accounting for correlated noise was the wavelet approach
of Carter \& Winn (2010).  This approach allowed for a power-law spectrum for the
noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate description of stellar variability.  Their approach runs in
$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
particular type of noise overpredicts the power spectrum at low frequency compared
to the granulation power spectrum.  It also cannot describe quasi-periodic noise.

>>>>

There are three primary applications we envision for the \genrp formalism: 1). modeling
the variability of an astrophysical system to infer it properties; 2).
accounting for astrophysical variability as a source of noise when trying to detect 
additional phenomena; 3). interpolating or extrapolat variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of quasars, measuring the asteroseismic variations of stars,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in 
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

\subsection{Application to stellar variability, transiting exoplanets}

Our background is in studying transiting exoplanets, a field which has only recently begun
to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
One promising early attempt at accounting for correlated noise was the wavelet approach
of Carter \& Winn (2010).  Their technique allowed for a power-law spectrum for the
noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.  Their approach runs in
$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
particular type of noise overpredicts the power spectrum at low frequency compared
to the granulation power spectrum.  It also cannot describe quasi-periodic noise.

Further progress was made by parameterizing the covariance matrix with simple,
analytic functions that describe the autocorrelation function.  The parameters
of these functions can then be optimized to best match the observed variability
pattern of the residuals (after subtracting a transit model).  The disadvantage of
this approach is that the functions used are chosen somewhat arbitrarily (e.g.\
the exponential-squared function), and the Cholesky decomposition of the covariance
matrix for computing the likelihood takes $\cal{O}(N^3)$ operations which prohibits 
application of this technique to large datasets.

Given the drawbacks of these approaches, the \genrp formalism allows both a fast
computation of the likelihood in $\cal{O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \genrp
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.
